---
title: "Predicting Tuition Costs"
author: "Shaun Sandbloom - srs341@georgetown.edu"
date: "12/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

\begin{center}
\Huge{PPOL 670 | Assignment 6}

\Huge{Supervised Machine Learning}
\end{center}

\vspace{0.1in}

<style>
    pre {
        border: 0;
    }
</style>

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(tidymodels)
library(jsonlite)
library(httr)
library(rscorecard)
library(sf)
library(tigris)
library(tidycensus)
library(usmap)
library(recipes)
library(rpart)
```

## Introduction

* The student debt crisis is a frequent topic of discussion in policy environments.
* There are a wide variety of factors that contribute to the problem.
* But debt starts with one key factor - **cost**.


#### Meaning of Education

* "Education" can mean something different to everyone. 
* Formal, structured post-secondary education occurs in a variety of formats.
* What appears to stand out is what it takes to earn a **bachelor's degree** from a **four-year academic program** commonly offered by universities.
* This analysis explores the characteristics of both **public** and **private (non-profit)** universities that award bachelor's degrees for completing a **four-year** academic program.

## Data Source

#### [The U.S. Department of Education College Scorecard API](https://collegescorecard.ed.gov/data/documentation/)

- Data documentation can be found [here](https://collegescorecard.ed.gov/data/documentation/)
- The API data dictionary can be found [here](https://collegescorecard.ed.gov/assets/FullDataDocumentation.pdf)

## Extraction, Transformation, and Loading (ETL)

The College Scorecard API uses a parameter-based query structure. Additionally, the API will only return 100 rows of data at a time. This means queries with large-volume requests will need to be iterated over multiple requests and aggregated into a larger data frame.

To simplify queries to the API, I first created a user-defined function to format the desired attributes appropriately for the request URL.

```{r, eval = FALSE}
query_scorecard <- function(endpoint, filters, fields, per_page, page){
  api_path <- "https://api.data.gov/ed/collegescorecard/v1/"
  query <- paste0(
    api_path, 
    endpoint,
    ".json?",
    filters, 
    "&fields=", 
    paste(fields, collapse = ","),
    "&per_page=",
    per_page,
    "&page=",
    page,
    "&api_key=", 
    api_key
    )
  print(query)
  print("Querying...")
  response <- GET(query)
  if(http_error(response)) {
    print("Query failed.")
    print(http_status(response)$message)
  } else {
    print("Query confirmed.")
    print(paste("Data type:", http_type(response)))

    content <- content(response, as = "text")
    results <- fromJSON(content)
    final <- tibble(results$results)
  }
}
```

#### Variables collected

The following data points are the result of pulling data from the **College Scorecard API** from 2000-2019 and extensive transformation. The Department of Education publishes an [Excel spreadsheet](https://collegescorecard.ed.gov/assets/CollegeScorecardDataDictionary.xlsx) with over 2000 variables on academic institutions. 

| Variable | Description |
| :------- | :---------- |
| `id` | A school's IPED code |
| `name` | A school's name |
| `city` | The city the school is located |
| `state` | The state the school is located |
| `ownership` | A value denoting if a school is public or private non-profit |
| `tuition_type` | Distinguishes between in-state and out-of-state rates |
| `tuition_rate` | The financial cost of tuition for an academic and/or program year |
| `location.lat` | A school's latitudinal coordinate |
| `location.lon` | A school's longitudinal coordinate |
| `size` | Categorical description of the the school's size |
| `size_details` | Categorical description of the student body |
| `locale_type` | Categorical description of the school's geographic setting |
| `locale_size` | Categorical description of the size of the geographical setting |

## Exploratory Data Analysis (EDA)

#### How have tuition rates evolved over the last 20 years?

```{r, echo = FALSE}
territories <- c("AS", "FM", "GU", "MH", "MP", "PR", "PR", "PW", "VI")

master_final <- read_csv("data/master_final.csv") %>% 
  select(-communication:-military) %>% 
  filter(tuition_rate > 1000,
         !state %in% territories) %>% 
  mutate(ownership = as.factor(ownership),
         size = as.factor(size),
         size_details = as.factor(size_details),
         locale_type = as.factor(locale_type),
         locale_size = as.factor(locale_size),
         tuition_type = as.factor(tuition_type))

master_final %>% 
  group_by(year, ownership) %>% 
  summarize(total = sum(tuition_rate)) %>% 
  ggplot(aes(x = year, y = total, fill = ownership)) +
  geom_col(width=.5, position = "dodge") +
  theme_minimal() +
  labs(title = "Total Tuition Costs from 2000 to 2019",
       y = "Costs (in dollars)") +
  scale_y_continuous(labels = comma)
```

#### What's meaningful about this view

* Annual tuition costs appear to be steadily rising across both public and private institutions.
* As a whole, the cost of private higher-education appears to be significantly more expensive.

#### What this view doesn't answer
* Calculating total costs on an annual basis overlooks the number of schools in each category...
* Are there simply more private schools than public schools?

```{r, echo = FALSE}
master_final %>% 
  group_by(year, ownership) %>% 
  summarize(total = n()) %>% 
  ggplot(aes(x = year, y = total, fill = ownership)) +
  geom_col(width=.5, position = "dodge") +
  theme_minimal() +
  labs(title = "Total Number of Schools from 2000 to 2019",
       y = "# of Schools") +
  scale_y_continuous(labels = comma)
```


#### Comparing tution rates from state-to-state
```{r, echo = FALSE}
territories <- c("AS", "FM", "GU", "MH", "MP", "PR", "PR", "PW", "VI")

pub_rates_in <- master_final %>% 
  filter(!state %in% territories,
         year == 2019,
         ownership == "public",
         tuition_type == "tuition_in_state") %>% 
  group_by(state, tuition_type) %>% 
  summarize(total = mean(tuition_rate))

plot_usmap(data = pub_rates_in, values = "total", color = "white", exclude = "AL") +
  scale_fill_continuous(low = 'green', 
                        high = 'red',
                        label = scales::comma,
                        limits = c(0,50000),
                        name = "Dollars ($)") +
  theme(legend.position = "right") +
  labs(title = "Average In-State Public Tuition Rates (2019)")

pub_rates_out <- master_final %>% 
  filter(!state %in% territories,
         year == 2019,
         ownership == "public",
         tuition_type == "tuition_out_state") %>% 
  group_by(state, tuition_type) %>% 
  summarize(total = mean(tuition_rate))

plot_usmap(data = pub_rates_out, values = "total", color = "white", exclude = "AL") +
  scale_fill_continuous(low = 'green', 
                        high = 'red',
                        label = scales::comma,
                        limits = c(0,50000),
                        name = "Dollars($)") +
  theme(legend.position = "right") +
  labs(title = "Average Out-of_State Public Tuition Rates (2019)")

priv_rates <- master_final %>% 
  filter(!state %in% territories,
         year == 2019,
         ownership == "private") %>% 
  group_by(state) %>% 
  summarize(total = mean(tuition_rate))

plot_usmap(data = priv_rates, values = "total", color = "white", exclude = c("AK", "WY")) +
  scale_fill_continuous(low = 'green', 
                        high = 'red',
                        label = scales::comma,
                        limits = c(0,50000),
                        name = "Dollars ($)") +
  theme(legend.position = "right") +
  labs(title = "Average Private Tuition Rates (2019)")
```

#### Wanna get away?

* There's a clear financial advantage to attending a four-year public institution in your home state.
* Students who choose to attend an out-of-state public institution might consider nearby private alternatives, **depending on the state**.

### One last exploration into the evolution to tuition costs...

#### Most Expensive and Least Expensive Schools: 2019
```{r, echo = FALSE}
master_final %>% 
  filter(year == 2019) %>% 
  select(name, city, state, ownership, tuition_rate) %>% 
  group_by(name, city, state, ownership) %>% 
  summarize(cost = max(tuition_rate)) %>% 
  arrange(desc(cost)) %>% 
  head() %>% 
  knitr::kable()

master_final %>% 
  filter(year == 2019) %>% 
  select(name, city, state, ownership, tuition_rate) %>% 
  group_by(name, city, state, ownership) %>% 
  summarize(cost = max(tuition_rate)) %>% 
  arrange(desc(cost)) %>% 
  tail() %>% 
  knitr::kable()
```

#### Most Expensive and Least Expensive Schools: 2000
```{r, echo = FALSE}
master_final %>% 
  filter(year == 2000) %>% 
  select(name, city, state, ownership, tuition_rate) %>% 
  group_by(name, city, state, ownership) %>% 
  summarize(cost = max(tuition_rate)) %>% 
  arrange(desc(cost)) %>% 
  head() %>% 
  knitr::kable()

master_final %>% 
  filter(year == 2000) %>% 
  select(name, city, state, ownership, tuition_rate) %>% 
  group_by(name, city, state, ownership) %>% 
  summarize(cost = max(tuition_rate)) %>% 
  arrange(desc(cost)) %>% 
  tail() %>% 
  knitr::kable()
```

## Continuing the discussion...

Future analyses may offer new insights on these data, by exploring predictive models and evaluating which of the variables above have a significant impact on predicting future tuition costs.